---
title: "Exercise 13"
subtitle: "Applied Statistics 2021, IT University of Copenhagen"
output:
  pdf_document: default
documentclass: article
classoption: a4paper
references:
- id: Dekking2010
  type: book
  title: "A Modern Introduction to Probability and Statistics: Understanding why and how"
  author:
  - family: Dekking
    given: F. M.
  - family: Kraaikamp
    given: C.
  - family: Lopuhaä
    given: H. P.
  - family: Meester
    given: L. E.
  issued:
    year: 2010
  publisher: Springer-Verlag
- id: Verzani2014
  type: book
  title: "Using R for Introductory Statistics"
  author:
  - family: Verzani
    given: John
  issued:
    year: 2014
  publisher: CRC Press
editor_options: 
  markdown: 
    wrap: 72
---

# Preparation

-   Read pages 283--284, 305--307, 321--325 from @Verzani2014.

# Exercises

```{r, message=FALSE, include=FALSE}
require(UsingR) 
library(magrittr)
library(ggplot2)
```

```{r}
area_poly <- function(cur, cutoff=c(-1, 1), side=c(1,-1), col = "grey", border=NA, ...)
{
    # right cutoff
    pos <- min(which(cur$x > cutoff[2]))
    end <- length(cur$x)

    polygon(x=c(cur$x[end:pos], cur$x[pos], cur$x[end]),
          y=c(cur$y[end:pos], 0, 0), col=col, border=border, ...)
  
    # left cutoff
    pos <- max(which(cur$x < cutoff[1]))
    end <- 1
    
    polygon(x=c(cur$x[end:pos], cur$x[pos], cur$x[end]),
          y=c(cur$y[end:pos], 0, 0), col=col, border=border, ...)
}
```

## 1. One sample $t$-test (T)

We perform a t-test for the null hypothesis $H_0 : \mu = 10$ by means of
a dataset consisting of $n=16$ elements with sample mean 11 and sample
variance 4. We use significance level 0.05.

a.  Should we reject the null hypothesis in favor of $H_1: \mu \neq 10$?

b.  What if we test against $H_1: \mu > 10$?

------------------------------------------------------------------------

Assuming normality of the data set (not explicitly mentioned in the task
description, but since theoretical exercise, assume to not perform
bootstrapping), we can compute the critical region for the test
statistic

$$
T = \frac{\bar{X}_n-\mu_0}{S_n/\sqrt{n}}
$$

from the fact that this distribution follows a *t-distribution* in
*m=n-1* degrees of freedom. The critical bounds for the alternative
hypothesis $H_1: \mu \ne 10$ (two-sided hypothesis testing), are
therefore given as the critical regions for the studentised mean
distribution:

```{r}
sample.mean <-  11
sample.var <- 4
n <- 16

t <- ((sample.mean - 10) / (sqrt(sample.var) / sqrt(16))) %>% print() # under null hypothesis
```

```{r}
alpha <- 0.05
critical.regions <- c(qt(alpha/2, df=n-1), qt(alpha/2, df=n-1, lower.tail = F))
print(critical.regions)

xs <- seq(- 5, 5, by = 0.01)
cc <- curve(dt(x, df=n-1), from=-5, to=5, n=100, type = "l", main = "Student's T-distribution PDF (with critical regions for a=0.05)", las=1)
area_poly(cc, cutoff = critical.regions, side=1, col = "red", density = 10)
abline(v=t, col='red')
```

Since our tests statistic $t$ is outside of the critical regions, we
assume the null hypothesis to be true until we found significant
evidence against it.

b\.

## 2. Easter Eggs (T)

Assume that you got six similar Easter eggs with 20g of chocolate
reported in each. After taking one more lecture in Applied Statistics,
you want to further investigate whether it is plausible that the eggs
really contain 20g chocolate or if the egg producer is cheating. You
weight the eggs and obtain the following six observations for the
chocolate weight:

|       Chocolate contents (g)       |
|:----------------------------------:|
| 20.1, 19.1, 18.2, 20.2, 19.6, 19.1 |

You may assume that you measurement is a realization of a random sample
from a normal distribution $N(\mu, \sigma^2)$, where $\mu$ represents
the true average contents.

(a) Formulate the appropriate null hypothesis and alternative
    hypothesis.

(b) Which test is appropriate for testing the hypothesis? Explain why.

(c) Compute the value of the test statistic and report your conclusion
    at significance level $\alpha = 0.05$.

(d) Compute the corresponding left tail $p$-value. Is it likely to
    observe these measurements under the null hypothesis?

a\.

$$
H_0: \mu = 20 \\
H_1: \mu \ne 20 or H_1: \mu \lt 20
$$

b\. In this specific scenario we would test with a one-sample *t-test*
to evaluate the hypotheses at hand, since we only deal with one sample,
namely the measurement of weights of the chocolate and can assume
normality of the data, such that the random variable $X$ representing
the weight of chocolate in a single easter egg, normally distributed
with $N(\mu, \sigma^2)$, where $\mu$ is the true mean average weight of
the chocolate.

c\. The test statistic $T$ to be used is the studentised mean, that
describes the standardised difference of the sample mean from the
hypothesised mean:

$$
T = \frac{\bar{X}_n-\mu_0}{S_n/\sqrt{n}} = 0.3048679
$$

```{r}
data <- c(20.1, 19.1, 18.2, 20.2, 19.6, 19.1)
studentised.mean <- ((mean(data) - 20) / (sd(data) / sqrt(length(data)))) %>% print()
```

Since the test statistic of the studentised mean follows a
t-distribution in n-1 degrees of freedom, we can either compute the
critical regions and observe whether or not our test statistic falls
into them or compute the p-value. For the alternative hypothesis, that
the true mean is not equal to 20 (not specifying in which direction it
deviates), we can compute the critical regions as follows:

```{r}
n <- length(data)
alpha <- 0.05
critical.regions <- c(qt(alpha/2, df=n-1), qt(alpha/2, df=n-1, lower.tail = F))  %>% print()
```

d\. The left-tail p-value corresponds to the probability that we get at
an as small value for the studentised mean, while assuming the null
hypothesis to be true.

$$
p-val = P(T \le -2.022734) = 0.04951219 
$$

```{r}
pt(-2.022734, df=n-1)
```

We would reject the null hypothesis in favour of the alternative
hypothesis. Furthermore we can conclude that we would assume the true
expectation of the weight of the chocolate in an easter egg to be lower
than 20.

## 3. Two-sample $t$-test (T)

The data in Table 28.3 (pp. 425, @Dekking2010) represents salaries (in
pounds Sterling) in 72 randomly selected advertisements in The Guardian
(April 6, 1992). When the range was given in the advertisement, the
midpoint of the range is reproduced in the table. The data are salaries
coreresponding to two kinds of occupations ($n=m=72$): (1) Creative,
media, and marketing and (2) education. The sample mean and sample
variance of the two datasets are, respectively:

$$
\begin{equation*}
\begin{split}
    (1)\ \bar{x}_{72} = 17 410 \textrm{ and } s_x^2= 41 258 741,\\
    (2)\ \bar{y}_{72} = 19 818 \textrm{ and } s_y^2= 50 744 521,
\end{split}
\end{equation*}
$$

Suppose that the datasets are modeled as realizations of normal
distributions with expectations $\mu_1$ and $\mu_2$, which represent the
salaries for occupations (1) and (2).

a.  Test the null hypothesis that the salary for both occupations is the
    same at level $\alpha = 0.05$ under the assumption of equal
    variances. Formulate the proper null and alternative hypotheses,
    compute the value of the test statistic, and report your conclusion.

b.  Do the same without the assumption of equal variances.

c.  As a comparison, one carries out an empirical bootstrap simulation
    for the nonpooled studentized mean difference. The bootstrap
    approximations for the critical values are $c_l^* = -2.004$ and
    $c_u^*=2.133$. Report your conclusions about the salaries on the
    basis of the bootstrap results.

## 4. Significance test for the mean (R)

The United States Department of Energy conducts weekly phone surveys on
the price of gasoline sold in the United States. Suppose one week the
sample average was \$4.03, the sample standard deviation was \$0.42, and
the sample size was 800. Perform a one-sided significance test of
$H_0:\mu=4.00$ against the alternative $H_\mathrm{A}:\mu>4.00$.

```{r}
sample.mean <- 4.03
sample.sd <- 0.42
n <- 800
alpha <- 0.05

t <- ((sample.mean - 4) / (sample.sd / sqrt(800))) %>% print()
p.val <- pt(t, df=n-1, lower.tail=F) %>% print()
if (p.val < alpha) {
    print('Rejected Null Hypothesis in favor of Alternative Hypothesis')
}
critical.regions <- c(qt(alpha/2, df=n-1), qt(alpha/2, df=n-1, lower.tail = F)) %>% print()

```

## 5. Two-sample tests of centre (R)

The data set `normtemp (UsingR)` contains body measurements for 130
healthy, randomly selected individuals. The variable `temperature`
contains normal body temperature data and the variable \`gender contains
gender information, with male coded as 1 and female as 2. Can you assume
that the groups have similar standard deviation? Is the sample
difference across two groups statistically significant? Is the
conclusion the same if you made a different assumption of the standard
deviation?

```{r}
males <- subset(normtemp, gender == 1)
females <- subset(normtemp, gender == 2)
boxplot(males$temperature, females$temperature, labels=c('MALE', 'FEMALE'), 
        main='Boxplot of Body Temperature for Males and Females', 
        xlab='Gender (Left: Male; Right: Female', 
        ylab='Body Tempearture')
```

$$
H_0: \mu_M = \mu_F \\
H_0: \mu_M \ne \mu_F
$$

```{r}
# assuming equal standard deviation (bootstrapping with pooled-variance)
x <- males$temperature
y <- females$temperature

pooled.variance <- sqrt(var(x)/length(x) + var(y)/length(y))
studentised.mean <- (mean(x) - mean(y)) / pooled.variance

alpha <- 0.05; t.distribution.samples = c();

for (i in 1:10000) {
  bootstrap.x <- sample(x, length(x), replace=TRUE)
  bootstrap.y <- sample(y, length(y), replace=TRUE)
  bootstrap.pooled.variance <- sqrt(var(bootstrap.x)/length(x) + var(bootstrap.y)/length(y))
  bootstrap.studentised.mean <- ( (mean(bootstrap.x)-mean(bootstrap.y)) - (mean(x)-mean(y)) ) / bootstrap.pooled.variance
  t.distribution.samples <- c(t.distribution.samples, bootstrap.studentised.mean)
}
quantile(t.distribution.samples, probs=c(alpha/2, 1-alpha/2))
mean(t.distribution.samples < studentised.mean)
```

Our analysis shows evidence against the null-hypothesis. Given the
boxplots, this is also somewhat expected.

## 6. Bootstrapping in two-sample tests (R)

For the `babies (UsingR)` data set, the variable `age` contains the
recorded mom's age and `dage` contains the dad´s age for several
different cases in the sample. Do a significance test of the null
hypothesis of equal age against a one-sided alternative that dads are
older in the sample population. Use a non-normal model with
bootstrapping.

```{r}
dad.ages <- babies$dage
mum.ages <- babies$age

boxplot(dad.ages, mum.ages, labels=c('MALE', 'FEMALE'), 
        main='Boxplot of Body Temperature for Males and Females', 
        xlab='Gender (Left: Male; Right: Female', 
        ylab='Body Tempearture')
```

```{r}
# assuming non normality of the data (non-pooled studentised mean)
x <- dad.ages
y <- mum.ages

nonpooled.var <- sqrt( var(x)/length(x) + var(y)/length(y) )
nonpooled.studentised.mean <- (mean(x) - mean(y)) / nonpooled.var

alpha <- 0.05; t.distribution.samples = c();

for (i in 1:10000) {
  bootstrap.x <- sample(x, length(x), replace=TRUE)
  bootstrap.y <- sample(y, length(y), replace=TRUE)
  bootstrap.pooled.variance <- sqrt(var(bootstrap.x)/length(x) + var(bootstrap.y)/length(y))
  bootstrap.studentised.mean <- ( (mean(bootstrap.x)-mean(bootstrap.y)) - (mean(x)-mean(y)) ) / bootstrap.pooled.variance
  t.distribution.samples <- c(t.distribution.samples, bootstrap.studentised.mean)
}
quantile(t.distribution.samples, probs=c(alpha/2, 1-alpha/2))
mean(t.distribution.samples < studentised.mean)
```
