---
title: "Bootstrap"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bootstrapping the mean

We can use the empirical bootstrap to look at the distribution of our estimate of the mean. Like in the book, consider the Old Faithful dataset.

```{r}
eruptions <- faithful$eruptions
sample_mean <- mean(eruptions)
bootstrap_means <- c()
bootstrap_means_sample_mean_difference <- c()
bootstrap_means_sample_mean_deviation <- c()
for (i in 1:1000) {
    bootstrap_sample <- sample(eruptions, size = length(eruptions), replace = TRUE)
    bootstrap_means <- c(bootstrap_means, mean(bootstrap_sample))
    bootstrap_means_sample_mean_difference <- c(bootstrap_means_sample_mean_difference, mean(bootstrap_sample) - sample_mean )
    bootstrap_means_sample_mean_deviation <- c(bootstrap_means_sample_mean_deviation, abs(mean(bootstrap_sample) - sample_mean ))
}
plot(density(bootstrap_means))
abline(v=sample_mean, col='blue')
plot(density(bootstrap_means_sample_mean_difference))
abline(v=5/60, col='red')
abline(v=-5/60, col='red')
plot(density(bootstrap_means_sample_mean_deviation), xlim=c(0, max(bootstrap_means_sample_mean_deviation)))
abline(v=5/60, col='red')
```

We can use these bootstrapped means to bound the probability that the sample mean is away from the norm by more than, say, 5 seconds (note that the times in the dataset are in minutes, as opposed to the book, where they are in seconds).

```{r}
bootstrap_means_difference_distribution <- ecdf(bootstrap_means_sample_mean_difference)
probability_outside <- 1 - (bootstrap_means_difference_distribution(5/60) - bootstrap_means_difference_distribution(-5/60))
probability_outside

x <- ecdf(bootstrap_means_sample_mean_deviation)
p <- 1 - x(5/60)
```

# KS distance on the software dataset

```{r}
software<- c(0,0,0,2,4,6,8,9,10,10,10,12,15,15,16,21,22,24,26,30,30,31,33,36,44,50,55,58,65,68,
             75,77,79,81,88,91,97,100,108,108,
             112,113,114,115,120,122,129,134,138,143,
             148,160,176,180,193,193,197,227,232,233,
             236,242,245,255,261,263,281,290,296,300,
             300,325,330,357,365,369,371,379,386,422,
             445,446,447,452,457,482,529,529,543,600,
             648,670,700,707,724,729,748,790,810,816,
             828,843,860,865,868,875,943,948,983,990,
             1011,1045,1064,1071,1082,1146,1160,1222,1247,1351,
             1435,1461,1755,1783,1800,1864,1897,2323,2930,3110,
             3321,4116,5485,5509,6150)
plot(density(software))
plot(ecdf(software))
lambda_estimate = 1/mean(software)
curve(pexp(x, rate=lambda_estimate), col='red', add=T)
```

Is the exponential distribution with the estimated parameter really a good fit?

```{r}
ks_distance_exp_distribution <- function(bootstrapped_data, lambda) {
  empirical_distribution <- ecdf(bootstrapped_data)
  max(abs(empirical_distribution(bootstrapped_data) - pexp(bootstrapped_data, rate = lambda)))
}
ks_estimate <- ks_distance_exp_distribution(software, lambda_estimate)

bootstrap_ks <- c()
for (i in 1:1000) {
  bootstrap_sample <- rexp(length(software), lambda_estimate)
  bootstrap_lambda <- 1/mean(bootstrap_sample)
  bootstrap_ks <- c(bootstrap_ks, ks_distance_exp_distribution(bootstrap_sample, bootstrap_lambda))
}
plot(density(bootstrap_ks), xlim=c(0,0.2))
abline(v=ks_estimate, col='red')
```

Looks like it's not!
