---
title: 'My First R-Markdown'
author: 'Jonas-Mika Senghaas'
date: '10/02/2021'
# output: pdf_document
---

## Code cell
This is an example of a simple code cell. 
```{r}
2+2
```
R allows you to make more complicated expressions. You can also mute the input of the cell by adding the flag 'echo = FALSE' into the constructor of the code cell.
```{r}
sqrt(16) # this is the square root of 16
```

## Functions
```{r}
my.adder <- function(x, add.this = 1) {
  result <- x + add.this
  return(result)
}

my.adder(5,10)
```

## Flow Control 

### If-Elif-Else

```{r}
is_positive <- function(x) {
  if(x >= 0) {
    return(TRUE)
  } else {
    return(FALSE)
  }
}

is_positive(2)

```


### For-Loop
```{r}
vector <- c(1:10)
vector

for (i in vector) {
  print(i)
}
```


# Statistics in R
We will explore R's capabilities of statistics by exploring the famous iris dataset that is.

```{r}
iris_data <- data(iris)
head(iris)
```

### Summarise Data
```{r}
summary(iris)
```

### Get the mean
```{r}
calculate.mean <- function(x) {
  return(sum(x)/length(x))
}
# this however is already implemented in R in the following way
mean(iris$Sepal.Length)
calculate.mean(iris$Sepal.Length)
```

### Sample Variance and Standard Deviation
```{r}
calculate.variance <- function(x) {
  return(1/(length(x)-1) * sum((x-mean(x))^2)) # this is the formula for the variance
}
calculate.variance(iris$Sepal.Length) # variance
var(iris$Sepal.Length)

sd(iris$Sepal.Length) # standard deviatin

```

### Dealing with Missing Values
```{r}
#mean(iris$Sepal.Length, na_rm = TRUE)
max(iris$Sepal.Length, na_rm = TRUE)
#min(iris$Sepal.Length, na_rm = TRUE)
#median(iris$Sepal.Length, na_rm = TRUE)
#var(iris$Sepal.Length, na_rm = TRUE)
#sd(iris$Sepal.Length, na_rm = TRUE)
```

## Simulation

```{r}
#set_seed(42)

#this is an example of a coin tossing simulation
sample(
  x= c("H", "T"),
  prob = c(.5, .5),
  size = 10,
  replace = TRUE
)
```

## Probability Distributions
```{r}
rnorm(5) # get 5 samples from the gaussian elimination
```

## Plotting
```{r}
plot(iris$Petal.Length, iris$Petal.Width, col = iris$Species, main="Iris Data")

# legend("topleft", legend=level(iris$Species), )
```

```{r}
a.vec <- subset(iris, Species == "setosa")$Sepal.Length

hist(a.vec, 
     col = 'skyblue',
     breaks=50)
```

```{r}
samples <- rnorm(1000)
hist(samples, freq=FALSE, ylim = c(0, 0.45), main = "Histogram of Gaussian Random Numbers (1000 Samples)")

xs <- seq(-5, 5, by= 0.01)
probs = rnrom(xs)
lines(xs, probs, col='red')
```
Saving Figures
```{r}
require(dplyr)
png(file = 'pie.png')

data <- subset(starwars, species %in% c('Human', 'Droid', 'Wookiee'))

pie(table(data$Species), main = 'Pie Chart')

dev.off()
```



## Tex-Code
$\Sigma_{i=0}^{N}i^2$ 
